---
title: "Introduction to the 'optconerrf' package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(optconerrf)
```

# Purpose of the package

The R package `optconerrf` can be used to derive the optimal conditional error function proposed by Brannath & Bauer (2004), which minimises the expected second-stage sample size of a two-stage adaptive clinical trial design while controlling for the conditional power.
Furthermore, in case the specification of the optimal conditional error function leads to a non-monotone function, the package can be used to implement the monotone transformation detailed in Brannath & Dreher (2024).

This document provides a brief introduction to the central function of the package.
These are:

* `getDesignOptimalConditionalErrorFunction()` to create a design which facilitates the use of the remaining functions 
* `getOptimalConditionalErrorFunction()` to calculate the optimal conditional error
* `getMonotonisationConstants()` to ensure the optimal conditional error function is non-increasing
* `getExpectedSecondStageSampleSize()` to calculate the expected second-stage sample size for a specification of the optimal conditional error function

# Prerequisites to computing the optimal conditional error function
\label{section_ocef}

The optimal conditional error function is defined as:

$$
\alpha_2(p_1) = \psi \left( \frac{-e^{c_0}}{Q(p_1)}\right).
$$

The individual parameters and functions are the following:

* $p_1$ is the first-stage p-value.
* $\psi$ is a function, for which no closed form is available and its solution is found through the `uniroot()` function in `optconerrf`. A requirement for this solution to exist is that the conditional power $CP$ is restricted to values $0.0228=1-\Phi(2)\leq CP \leq \Phi(2) = 0.9772$.
* $c_0$ is the level constant, required for the optimal conditional error function to preserve the overall type I error rate.
* $Q(p_1)=l(p_1)/\vartheta_1^2$ is the ratio of the likelihood ratio of $p_1$ and an effect assumption $\vartheta_1$ for the conditional power $CP$. The likelihood ratio requires an assumption about the true treatment effect, denoted $\vartheta$.

Note that $\vartheta$ and $\vartheta_1$ are specified as non-centrality parameters rather than the corresponding mean differences $\Delta$ and $\Delta_1$.
The non-centrality parameter $\vartheta$ (respectively, $\vartheta_1$) is proportional to $\Delta$ (respectively, $\Delta_1$) and additionally takes into account the sample size $n_t$ (of the treatment group), allocation ratio $r=n_t/n_c$ between the treatment groups and variance $\sigma^2$.
It is calculated as:

$$
\vartheta = \frac{\Delta}{\sigma}\cdot\sqrt{\frac{n_t}{r+1}}.
$$

The non-centrality parameter $\vartheta$ has a convenient relationship to a normally distributed test statistic $Z$, namely $\vartheta=Z$.
It also has a one-to-one relationship to the corresponding p-value, i.e. $\vartheta = \Phi^{-1}(1-p)$.

The `optconerrf` package is based on a design object which contains the required specifications of the optimal conditional error functions.
This object is created by the function `getDesignOptimalConditionalErrorFunction()` and can be simply passed to other functions instead of specifying each argument separately.
The following section explains how to derive the design object.

# Creating the design object

For the creation of a design object, the following parameters are required and discussed in more detail below:

* Overall type I error rate and early stopping boundaries
* Conditional power
* Effect for likelihood ratio
* Effect for conditional power
* Optional: Minimum and maximum conditional error

In addition to these arguments, which must be specified by the user, the following fields are automatically calculated and added to the design object:

* Level constant.
* Monotonisation constants to ensure a non-increasing conditional error function.

## Overall type I error rate and early stopping boundaries

The overall type I error rate of the design, usually denoted by $\alpha$ is specified to `getDesignOptimalConditionalErrorFunction()` by the argument `alpha`.
Early stopping boundaries $\alpha_1$ and $\alpha_0$ are specified via the respective arguments `alpha1` and `alpha0`, with the stopping rules:

* Stop and reject $H_0$ after the first stage if $p_1 \leq \alpha_1$.
* Stop and fail to reject $H_0$ after the first stage if $p_1 > \alpha_0$.

Note that $\alpha_0$ is a binding futility boundary, i.e., it is not permitted to continue the trial in case $p_1>\alpha_0$.

## Conditional power

The conditional power to be achieved at a certain planning alternative is provided to `getDesignOptimalConditionalErrorFunction()` by the argument `conditionalPower`.


## Effect for likelihood ratio

The likelihood ratio of the first-stage p-value $p_1$ for a normally distributed test statistic with true effect $\vartheta$ according to Hung et al. (1997) is calculated as:

$$
l(p_1) = e^{\Phi^{-1}(1-p_1)\cdot \vartheta -\frac{\vartheta^2}{2}}.
$$

To calculate the likelihood ratio with a "fixed" effect $\vartheta=1$, the arguments `dist = "fixed"` and `ncpLR = 1` must be specified for `getDesignOptimalConditionalErrorFunction()`.


Selecting a single value for `ncpLR` may however provide an inefficient design under misspecification and instead, one may wish to use weighted alternatives for this parameter.

A discrete weighting of the parameter $\vartheta$ for the likelihood ratio is achieved by specifying multiple values for `ncpLR`.
The respective weights can be provided by `weights` and are assumed to be equal if not explicitly specified.
An exemplary specification could be `ncpLR = c(0, 0.5, 1)` and `weights = c(0.25, 0.25, 0.5)`.
Note that `dist = "fixed"` must still be specified in this case.

The weighting of values for $\vartheta$ may also be done via a distributional assumption, i.e., a prior density $u(\vartheta)$.
The general form of the likelihood ratio is then:

$$
l(p_1) = \int_{-\infty}^{\infty}e^{\Phi^{-1}(1-p_1)\vartheta - \vartheta ^2/2}u(\vartheta)d\vartheta.
$$

Currently, a normal, exponential and uniform prior density $u(\vartheta)$ for $\vartheta$ are implemented.

A normal prior is specified via `dist = "normal"` and the mean `ncpLR` and standard deviation `tauLR`.
The use of an exponential prior is possible by `dist = "exp"` and requires the specification of its mean `kap0` (corresponding to the inverse of the rate parameter $\lambda$).
A uniform prior can be used with `dist = "unif"` and `delMax` specifies the maximum of its support [0, `delMax`]. 
It is also possible to use the data-driven maximum likelihood via `dist = "maxlr"`.
In this approach, the effect size $\vartheta$ is estimated from the data and used in the likelihood ratio calculation, provided it is not smaller than 0.
In this case, no additional parameters must be specified.

## Effect for conditional power

The optimal conditional error function controls for the conditional power of a design at a point alternative `ncp1` or an interim estimate with minimum `ncp0`.
The mode of action is determined by specifying either (and never both) of the aforementioned arguments to `getDesignOptimalConditionalErrorFunction()`.
For the case of calculating the conditional power at a fixed effect, `ncp1` must be specified.
A side note is that the actual size of the fixed effect does not matter (provided it is not 0), as the level constant $c_0$ will account for it.

For the use of an interim estimate, the cut-off `ncp0` must be provided.

## Optional: Minimum and maximum conditional error

In certain scenarios, it may be feasible to restrict the values that the optimal conditional error function can take by a minimum and/or maximum.
For the function `getDesignOptimalConditionalErrorFunction()`, this can be achieved by specifying values for the arguments `minimumConditionalError` and `maximumConditionalError`.
The default values for the arguments are `minimumConditionalError = 0` and `maximumConditionalError = 1`, corresponding to no restriction.

## Calculating the level constant

The level constant, required for the optimal conditional error function to meet the level condition, is automatically calculated whenever a design is created using `getDesignOptimalConditionalErrorFunction()`.
This is done via the function `getLevelConstant()`, which extracts the required parameters from the specified design object.

## Calculating monotonisation constants

The use of a conditional error function which is increasing for any first-stage p-values is discouraged due to a type I error inflation in the case of a conservatively distributed $p_1$ and because the resulting power function of the trial may be decreasing in the true effect size (Koyama et al., 2005).
Because the framework of the optimal conditional error function may however lead to an increasing conditional error function, a non-increasing transformation may be required.
To retain optimality, the non-monotone optimal conditional error function must be set to constant values on certain intervals.
How to identify the constants and intervals is described in Brannath & Dreher (2024).

Whenever a design object is created using `getDesignOptimalCondtionalErrorFunction()`, the function `getMonotonisationConstants()` is internally called to identify the intervals and constants.
If the specified design produces an optimal conditional error function which already is non-increasing in the first-stage p-value, no monotonisation is necessary.
In this case, a message is printed during the creation of the design object and the field `monotonisationConstants` is an empty list.

# Using the design object

Once the design object is created using `getDesignOptimalCondtionalErrorFunction()`, it can be easily passed to other functions of `optconerrf`.
The design object is a list and it can be easily printed for an overview:

```{r, message = FALSE}
design <- getDesignOptimalConditionalErrorFunction(
  alpha = 0.025, alpha1 = 0.001, alpha0 = 0.5, conditionalPower = 0.9,
  ncp1 = 1, dist = "fixed", ncpLR = 1
)

print(design)
```

To access individual fields of the object, e.g., the overall type I error rate (`alpha`) or the level constant (`levelConstant`) the following code can be used:

```{r}
design$alpha
design$levelConstant
```

For an overview of all available fields, the `names()` function can be used:

```{r}
names(design)
```

Note that only either the field `ncp1` or `ncp0` can exist, depending on the specification of the effect for conditional power.

# Computing the optimal conditional error function

Values for the optimal conditional error function are calculated using the function `getOptimalConditionalError()`.
For this function, the p-value of the first stage (argument `firstStagePValue`) and a design object, created using `getDesignOptimalConditionalErrorFunction()`, must be provided.

For the specification above, the optimal conditional error for an exemplary first-stage p-value of `0.1` can be calculated as:

```{r}
getOptimalConditionalError(
  firstStagePValue = 0.1, design = design
)
```

Multiple first-stage p-values may also be provided as a vector.
For first-stage p-values with $p_1 \leq \alpha_1$ or $p_1 > \alpha_0$, a conditional error of 1 or 0 is returned, respectively.

```{r}
getOptimalConditionalError(
  firstStagePValue = c(0.0005, 0.01, 0.05, 0.5, 0.8),
  design = design
)
```

Plotting the optimal conditional error function for a grid $p_1 \in [0,1]$ can also be done for this design:

```{r, fig.dim=c(6,4)}
p <- seq(0, 1, 0.001)
optConErr <- getOptimalConditionalError(
  firstStagePValue = p, design = design
)

plot(p, optConErr, type = "l", xlab = "First-stage p-value", ylab = "Optimal Conditional Error")
```

# Calculation of expected second-stage sample size
\label{section_exp}

With the function `getExpectedSecondStageSampleSize()`, a calculation of the expected second-stage sample size that a specification of the optimal conditional error function yields can be calculated.
The calculation of the expected second-stage sample size $\mathbb{E}(n_{2,t})$ is performed as:

$$
\mathbb{E}(n_{2,t}) = \int_{\alpha_1}^{\alpha_0} \frac{\sigma^2\cdot(r+1) \cdot (\Phi^{-1}(1-\alpha_2(p_1)) + \Phi^{-1}(CP_{\Delta_1}))^2\cdot l(p_1)}{\Delta_1^2}dp_1.
$$

In this formula, $r$ specifies the allocation ratio between the second-stage sample size of the treatment group ($n_{2,t}$) and control group ($n_{2,c}$) and is defined as $r=n_{2,t}/n_{2,c}$.
The likelihood ratio $l(p_1)$ defines for which scenario to calculate the expected second-stage sample size., i.e., for an effect under $H_0$ (e.g., $\vartheta=0$), a planning alternative $\vartheta$ or somewhere in between, e.g. $\vartheta/2$.
The effect $\Delta_1$ provides the effect for which the conditional power $CP_{\Delta_1}$ is targeted.
Again, the effect sizes $\vartheta$ and $\Delta_1$ may be different.
For example, one may wish to investigate the expected sample size under the null hypothesis ($\vartheta=0$), but the (conditional) power can only be reasonably specified for an effect size $\Delta_1>0$.

The calculation of the expected second-stage sample size of the treatment group for the above design, assuming $\vartheta$ was correctly specified, can be calculated as:

```{r}
getExpectedSecondStageSampleSize(
  design = design, distDelta = "fixed", ncpLRDelta = 1,
  allocationRatio = 1, standardDeviation = 1, delta1 = 0.25
)
```


The specification of the likelihood ratio under which to calculate the expected second-stage sample size is given by `distDelta="fixed"` and the corresponding non-centrality parameter is specified by `ncpLRDelta = 1`.
The effect size for the conditional power is specified via `delta1 = 0.25`.
The standard deviation $\sigma$ and the allocation ratio $r$ may be explicitly specified, otherwise the default values `standardDeviation=1` and `allocationRatio=1` are used by `getExpectedSecondStageSampleSize()`.

The expected second-stage sample size of the above design under the null hypothesis (for `ncpLRDelta = 0`) can be calculated as:

```{r}
getExpectedSecondStageSampleSize(
  design = design, distDelta = "fixed", ncpLRDelta = 0, delta1 = 0.25
)
```

Note that the sample sizes calculated above are the sample sizes of the treatment group.

# References

Brannath, W. & Bauer, P. (2004). Optimal conditional error functions for the control of conditional power. Biometrics, 60 (3), 715–723. https://doi.org/10.1111/j.0006-341X.2004.00221.x

Brannath, W. & Dreher, M. (2024). Optimal monotone conditional error functions. https://arxiv.org/abs/2402.00814

Hung, H. M. J., O’Neill, R. T., Bauer, P. & Kohne, K. (1997). The behavior of the p-value when the alternative hypothesis is true. Biometrics. http://www.jstor.org/stable/2533093
