---
title: "Introduction to the 'optconerrf' package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.dim = c(9, 6)
)
options(rmarkdown.html_vignette.check_title = FALSE)
```

```{r setup}
library(optconerrf)
```

# Purpose of the package

The R package `optconerrf` can be used to derive the optimal conditional error function proposed by Brannath & Bauer (2004), which minimises the expected second-stage sample size of a two-stage adaptive clinical trial design while controlling for the conditional power.
The package implements a more general approach by using the second-stage *information* rather than the second-stage *sample-size* and introduces extensions to the original methodology.
In case the specification of the optimal conditional error function leads to a non-monotone function, the package can be used to implement the monotone transformation detailed in Brannath et al. (2024).

This document provides a brief introduction to the central functions of the package.
These are:

* `getDesignOptimalConditionalErrorFunction()` to create a design which facilitates the use of the remaining functions 
* `getOptimalConditionalErrorFunction()` to calculate the optimal conditional error for a given first-stage p-value
* `getSecondStageInformation()` to calculate the second-stage information required to reach the target conditional power for a given first-stage p-value
* `getExpectedSecondStageInformation()` to calculate the expected second-stage information for a specification of the optimal conditional error function under an effect assumption

# Prerequisites to computing the optimal conditional error function
\label{section_ocef}

In `optconerrf`, the tested set of hypotheses is always assumed to be:

$$
H_0: \Delta \leq 0 \enspace \text{vs.} \enspace H_1: \Delta>0,
$$
where $\Delta$ is the effect size of interest.

The optimal conditional error function is defined as:

$$
\alpha_2(p_1) = \psi \left( \frac{-e^{c_0}}{Q(p_1)}\right).
$$

The individual parameters and functions are the following:

* $p_1$ is the first-stage p-value.
* $\psi$ is the inverse of the function $\nu'(u) = -2\cdot(\Phi^{-1}(1-u)+\Phi^{-1}(CP))/\phi(\Phi^{-1}(1-u))$. For $\psi$, no closed form is available and its solution is found through the `uniroot()` function in `optconerrf`.
* $c_0$ is the level constant, required for the optimal conditional error function to preserve the overall type I error rate.
* $Q(p_1)=l(p_1)/\vartheta_1^2$ is the ratio of the likelihood ratio $l(p_1)$ of $p_1$ and an effect assumption $\vartheta_1$ at which the conditional power $CP$ should be reached. The likelihood ratio requires an assumption about the true treatment effect, denoted $\vartheta$. This is the effect assumption under which the second-stage information is minimised.

Note that $\vartheta$ and $\vartheta_1$ denote non-centrality parameters rather than the corresponding mean differences $\Delta$ and $\Delta_1$.
The non-centrality parameter $\vartheta$ (respectively, $\vartheta_1$) is proportional to $\Delta$ (respectively, $\Delta_1$) and additionally takes into account the first-stage information $I_1$ of the trial.
It is calculated as:

$$
\vartheta = \Delta\cdot \sqrt{I_1}.
$$

In `optconerrf`, the treatment effect for the conditional power can be specified either as the mean difference `delta1` or the non-centrality parameter `ncp1`.

The `optconerrf` package is based on a design object which contains the required specifications of the optimal conditional error functions.
This object is created by the function `getDesignOptimalConditionalErrorFunction()` and can be simply passed to other functions instead of specifying each argument separately.
The following section explains how to derive the design object.

# Creating the design object

For the creation of a design object, the following parameters are required and discussed in more detail below:

* Overall type I error rate and early stopping boundaries
* Effect for likelihood ratio
* Conditional power
* Effect for conditional power
* First-stage information
* Optional: Constraints providing minimum and maximum conditional error (or maximum and minimum second-stage information, respectively)

In addition to these arguments, which must be specified by the user, the following fields are automatically calculated and added to the design object:

* Level constant
* Monotonisation constants to ensure a non-increasing conditional error function (if necessary and not explicitly inhibited by the user)

## Overall type I error rate and early stopping boundaries

The overall type I error rate of the design, denoted $\alpha$, is specified to `getDesignOptimalConditionalErrorFunction()` by the argument `alpha`.
Early stopping boundaries $\alpha_1$ and $\alpha_0$ are specified via the respective arguments `alpha1` and `alpha0`, with the stopping rules:

* Stop and reject $H_0$ after the first stage if $p_1 \leq \alpha_1$.
* Stop and fail to reject $H_0$ after the first stage if $p_1 > \alpha_0$.

Note that $\alpha_0$ is a binding futility boundary, i.e., it is not permitted to continue the trial in case $p_1>\alpha_0$.
Designs with $\alpha_1 = 0$ and/or $\alpha_0 = 1$, i.e., designs with no early efficacy and/or futility stopping, are permitted.

## Effect for likelihood ratio

The likelihood ratio of the first-stage p-value $p_1$ for a normally distributed test statistic with true non-centrality parameter $\vartheta=\Delta \cdot \sqrt{I_1}$ according to Hung et al. (1997) is calculated as:

$$
l(p_1) = e^{\Phi^{-1}(1-p_1)\cdot \vartheta -\frac{\vartheta^2}{2}}.
$$

To use a fixed effect for the likelihood ratio in the specification of the optimal conditional error function, `likelihoodRatioDistribution="fixed"` must be specified to `getDesignOptimalConditionalErrorFunction()`. In addition, the effect size $\Delta$ must be provided on the **mean difference scale** via the argument `deltaLR.`
The first-stage information $I_1$ is extracted from the design object and then used together with $\Delta$ to calculate the non-centrality parameter $\vartheta$.

Selecting a single value for `deltaLR` (and consequently, $\vartheta$) may however provide an inefficient design under misspecification and instead, one may wish to use weighted alternatives for this parameter.

A discrete weighting of the parameter $\vartheta$ for the likelihood ratio is achieved by specifying multiple values for $\Delta$ via `deltaLR`.
The respective weights can be provided by `weightsDeltaLR` and are assumed to be equal if not explicitly specified.
An exemplary specification could be `deltaLR = c(0, 0.5, 1)` and `weightsDeltaLR = c(0.25, 0.25, 0.5)`.
Note that `likelihoodRatioDistribution = "fixed"` must still be specified in this case.

The weighting of values for $\vartheta$ may also be done via a distributional assumption, i.e., a prior density $u(\vartheta)$.
The general form of the likelihood ratio is then:

$$
l(p_1) = \int_{-\infty}^{\infty}e^{\Phi^{-1}(1-p_1)\vartheta - \vartheta ^2/2}u(\vartheta)d\vartheta.
$$

Currently, a normal, exponential and uniform prior density $u(\vartheta)$ for $\vartheta$ are implemented.

A normal prior is specified via `likelihoodRatioDistribution = "normal"` and the mean `deltaLR` and standard deviation `tauLR`.
The use of an exponential prior is possible by `likelihoodRatioDistribution = "exp"` and requires the specification of its mean `kappaLR` (corresponding to the inverse of the rate parameter $\lambda$).
A uniform prior can be used with `likelihoodRatioDistribution = "unif"` and `deltaMaxLR` specifies the maximum of its support [0, `deltaMaxLR`]. 
It is also possible to use the data-driven maximum likelihood via `likelihoodRatioDistribution = "maxlr"`.
In this approach, the effect size $\vartheta$ is estimated from the data and used in the likelihood ratio calculation, provided it is not smaller than 0.
In this case, no additional parameters must be specified.

The arguments `deltaLR`, `tauLR`, `kappaLR` and `deltaMaxLR` must be specified on the mean difference scale, i.e., without accounting for the first-stage information.

## Conditional power

The conditional power to be achieved at a certain planning alternative is provided to `getDesignOptimalConditionalErrorFunction()` by the argument `conditionalPower`.
Instead of a fixed value, a user-defined conditional power function may also be provided by the argument `conditionalPowerFunction`.
This function should be specified in a way that it calculates the target conditional power from the first-stage p-value.
Internally, the conditional power function is provided with the first-stage p-value as its unnamed first and only argument.
Furthermore, it should generally be non-increasing in the first-stage p-value.
This is checked when initialising the design object (by using a grid of values) and a warning is displayed if the conditional power function is increasing.
For this purpose, the conditional power function should be able to handle vectors in its argument.

## Effect for conditional power

The optimal conditional error function controls for the conditional power of a design at a fixed point alternative or at an interim estimate of the treatment effect.
To use a fixed point alternative, `useInterimEstimate=FALSE` must be specified as well as a value for the point alternative via `delta1` (mean difference scale) or `ncp1` (non-centrality parameter scale).
Apart from a fixed effect, the treatment effect at which the conditional power should be achieved can also be estimated from the first-stage data.
In this case, it is necessary that a lower limit for this interim estimate is specified, as treatment effects $\Delta_1 \leq 0$ (respectively, $\vartheta_1 \leq 0$) belong to the null hypothesis and cannot be powered for. If `useInterimEstimate=TRUE` is specified, this lower limit must be provided via `delta1Min` or `ncp1Min`.
Furthermore, one may wish to deselect overly large interim treatment effect estimates by imposing an upper limit on the estimator.
This can be done by using the arguments `delta1Max` or `ncp1Max`, but is not required.

Note that the used effect specification is always used in the functions that calculate the second-stage information, i.e., `getSecondStageInformation()` and `getExpectedSecondStageInformation()`.

## First-stage information

The first-stage information of the trial design must be specified to allow for calculations between the mean difference and non-centrality parameter scale.
It is provided to the design object via `firstStageInformation`.

Listed below are some examples for the calculation between information ($I_1$) and sample size:

One-sample z-test with $n$ total patients: $I_1 = \frac{n}{\sigma^2}$, where $\sigma^2$ is the variance of an individual observation

Balanced two-sample z-test with $n_1$ patients per group: $I_1 = \frac{1}{2}\cdot\frac{n_1}{\sigma^2}$, where $\sigma^2$ is the common variance

General two-sample z-test with $n_1$, $n_2$ patients per group: $I_1 = 1/(\frac{\sigma_1^2}{n_1}+\frac{\sigma_2^2}{n_2})$, where $\sigma_1^2$, $\sigma_2^2$ are the group-wise variances 

## Optional: Minimum and maximum conditional error

In certain scenarios, it may be feasible to restrict the values that the optimal conditional error function can take by a minimum and/or maximum.
For the function `getDesignOptimalConditionalErrorFunction()`, this can be achieved by specifying values for the arguments `minimumConditionalError` and `maximumConditionalError`.
The default values for the arguments are `minimumConditionalError = 0` and `maximumConditionalError = 1`, corresponding to no restrictions.
Alternatively, the second-stage information may be restricted by using the arguments `minimumSecondStageInformation` and `maximumSecondStageInformation`.
Internally, the restrictions are always applied on the conditional error scale and if they are provided on the information scale, a calculation of the respective values on the conditional error scale is performed.
If constraints are provided on the conditional error scale as well as on the information scale, the information constraints are preferred and the constraints on the conditional error scale are ignored.

## Calculating the level constant

The level constant, required for the optimal conditional error function to meet the level condition, is automatically calculated whenever a design is created using `getDesignOptimalConditionalErrorFunction()`.
The level constant ensures that the optimal conditional error function meets the overall type I error rate of the design and requires that:
$$
\alpha = \alpha_1 + \int_{\alpha_1}^{\alpha_0}\alpha_2(p_1)dp_1.
$$

This is done via the function `getLevelConstant()`, which extracts the required parameters from the specified design object.
This function uses the `uniroot()` function to search for the level constant, by default on an interval from 0 to 10.
In certain cases, in particular for design specifications with very large or very small non-centrality parameters, the level constant may lie outside of this interval.
For such designs, `uniroot()` produces an error. The search interval can be manually changed by tempering with the arguments `levelConstantMinimum` and `levelConstantMaximum`.
Typically, designs with very large non-centrality parameters will require a smaller level constant (i.e., a reduction of `levelConstantMinimum`) and designs with very small non-centrality parameters will require a larger level constant (i.e., an increase of `levelConstantMaximum`).

Note that imposing inappropriate restrictions on the optimal conditional error function via `minimumConditionalError` and `maximumConditionalError` or `maximumSecondStageInformation` and `minimumSecondStageInformation` may make it impossible to calculate a level constant which fulfills the level condition.

## Calculating monotonisation constants

The use of a conditional error function which is increasing for any first-stage p-values is discouraged due to a type I error inflation in the case of a conservatively distributed $p_1$ and because the resulting power function of the trial may be decreasing in the true effect size (Koyama et al., 2005).
Because the framework of the optimal conditional error function may however lead to an increasing conditional error function for certain specifications, a non-increasing transformation may be required.
To retain optimality, the non-monotone optimal conditional error function must be set to constant values on certain intervals.
How to identify the constants and intervals is described in detail in Brannath et al. (2024).

Whenever a design object is created using `getDesignOptimalCondtionalErrorFunction()`, the function `getMonotonisationConstants()` is internally called to identify the intervals and constants.
If the specified design produces an optimal conditional error function which already is non-increasing in the first-stage p-value, no monotonisation is necessary.
For illustration purposes, `optconerrf` also allows the user to skip the monotonisation process by setting `enforceMonotonicity=FALSE`. 
In this case, the resulting conditional error function may be increasing for some p-values. 
Note that generally, such a design should not be used for practical purposes.

# Using the design object

As an example, we consider an adaptive design with overall type I error rate `alpha=0.025`, early rejection and futility boundaries `alpha1=0.0154` and `alpha0=0.5` with a target conditional power `conditionalPower=0.9` at a fixed alternative `delta1=0.25` (note that `useInterimEstimate=FALSE`).
We assume a fixed effect of `deltaLR=0.25` for the calculation of the likelihood ratio (`likelihoodRatioDistribution="fixed"`) and `firstStageInformation=50`.
Note that this design approximates the original trial design used in Brannath & Bauer (2004) and achieves an overall power of 90%.

```{r, message = FALSE, results = "markup"}
design <- getDesignOptimalConditionalErrorFunction(
  alpha = 0.025, alpha1 = 0.0154, alpha0 = 0.5, conditionalPower = 0.9,
  delta1 = 0.25, likelihoodRatioDistribution = "fixed", deltaLR = 0.25,
  firstStageInformation = 50, useInterimEstimate = FALSE
)
```

Once the design object is created using `getDesignOptimalCondtionalErrorFunction()`, it can be easily passed to other functions of `optconerrf`.
The design object is an instance of the class `TrialDesignOptimalConditionalError` and offers implementations of the generic functions `print()` and `plot()`:

```{r, message = FALSE, results = "markup"}
# Comprehensive design output
print(design)
```

```{r, fig.dim = c(6,4)}
plot(design) # Plot of the optimal conditional error function
plot(design, type = 2) # Plot of the second-stage information
```

To access individual fields of the object, e.g., the overall type I error rate (`alpha`) or the level constant (`levelConstant`) the following code can be used:

```{r} 
design$alpha
design$levelConstant
print(design$levelConstant, digits = 12) # For more precision
```

# Computing the optimal conditional error function and second-stage information

Values for the optimal conditional error function are calculated using the function `getOptimalConditionalError()`.
For this function, the p-value of the first stage (argument `firstStagePValue`) and a design object, created using `getDesignOptimalConditionalErrorFunction()`, must be provided.

For the specification above, the optimal conditional error for an exemplary first-stage p-value of `0.1` can be calculated as:

```{r}
getOptimalConditionalError(
  firstStagePValue = 0.1, design = design
)
```

Multiple first-stage p-values may also be provided as a vector.
For first-stage p-values with $p_1 \leq \alpha_1$ or $p_1 > \alpha_0$, a conditional error of 1 or 0 is returned, respectively.

```{r}
getOptimalConditionalError(
  firstStagePValue = c(0.0005, 0.1, 0.05, 0.5, 0.8),
  design = design
)
```

The calculation of the second-stage information is achieved in a similar fashion, using `getSecondStageInformation()`:

```{r}
getSecondStageInformation(
  firstStagePValue = 0.1, design = design
)

getSecondStageInformation(
  firstStagePValue = c(0.0005, 0.1, 0.05, 0.5, 0.8),
  design = design
)
```

Note that the second-stage information for first-stage p-values with $p_1 \leq \alpha_1$ or $p_1 > \alpha_0$ is always 0, as the trial is stopped early for such results.

# Calculation of expected second-stage information
\label{section_exp}

With the function `getExpectedSecondStageInformation()`, a calculation of the expected second-stage information that a specification of the optimal conditional error function yields can be calculated.
The calculation of the expected second-stage information $\mathbb{E}(I_2)$ is performed as:

$$
\mathbb{E}(I_2) = \int_{\alpha_1}^{\alpha_0} \frac{ (\Phi^{-1}(1-\alpha_2(p_1)) + \Phi^{-1}(CP_{\Delta_1}))^2\cdot l(p_1)}{\Delta_1^2}dp_1.
$$
The effect $\Delta_1$ provides the effect for which the conditional power $CP_{\Delta_1}$ is targeted.
$\Delta_1$ is extracted from the design object and may depend on the interim data.

The likelihood ratio $l(p_1)$ in this case represents the scenario under which the performance of a specific design is to be evaluated.
For example, one may have specified a design assuming $\Delta=0.25$ as a fixed effect for the likelihood ratio (as above), but now wishes to investigate the performance of this design in case $\Delta=0$, i.e., if the null hypothesis is true.
This can be achieved by the following code:

```{r}
getExpectedSecondStageInformation(
  design = design, likelihoodRatioDistribution = "fixed", deltaLR = 0
)
```

The specification of the likelihood ratio under which to calculate the expected second-stage information is given by `likelihoodRatioDistribution = "fixed"` and the corresponding treatment effect is specified by `deltaLR = 0`.

Alternatively, one may calculate the expected second-stage information of the design if it was correctly specified, i.e., for $\Delta=0.25$:

```{r}
getExpectedSecondStageInformation(
  design = design, likelihoodRatioDistribution = "fixed", deltaLR = 0.25
)
```

A shortcut for the calculation of the expected second-stage information under correct specification is also available by omitting the arguments `likelihoodRatioDistribution` and `deltaLR`, in which case the likelihood ratio specification given in the design is extracted and used for the calculation.

```{r}
getExpectedSecondStageInformation(
  design = design
)
```

Multiplying the respective information values by a factor of 2 (approximately) yields the original results of Brannath & Bauer (2004).

# References

Brannath, W. & Bauer, P. (2004). Optimal conditional error functions for the control of conditional power. Biometrics, 60 (3), 715–723. https://doi.org/10.1111/j.0006-341X.2004.00221.x

Brannath, W., Dreher, M., Scharpenberg, M. (2024). Optimal monotone conditional error functions. https://arxiv.org/abs/2402.00814

Hung, H. M. J., O’Neill, R. T., Bauer, P. & Kohne, K. (1997). The behavior of the p-value when the alternative hypothesis is true. Biometrics. http://www.jstor.org/stable/2533093
